%!TEX TS-program = lualatex
%!TEX encoding = UTF-8 Unicode

\documentclass[letterpaper]{tufte-handout}

%\geometry{showframe} % display margins for debugging page layout

\usepackage{fontspec}
\def\mainfont{Linux Libertine O}
\setmainfont[Ligatures={Common,TeX}, Contextuals={NoAlternate}, BoldFont={* Bold}, ItalicFont={* Italic}, Numbers={OldStyle}]{\mainfont}
\setsansfont[Scale=MatchLowercase]{Linux Biolinum O} 
\usepackage{microtype}

\usepackage{graphicx} % allow embedded images
  \setkeys{Gin}{width=\linewidth,totalheight=\textheight,keepaspectratio}
  \graphicspath{{img/}} % set of paths to search for images
\usepackage{amsmath}  % extended mathematics
\usepackage{booktabs} % book-quality tables
\usepackage{units}    % non-stacked fractions and better unit spacing
\usepackage{siunitx}
\usepackage{multicol} % multiple column layout facilities
\usepackage{microtype}   % filler text
\usepackage{hyperref}

\usepackage{enumitem}

%\usepackage{fancyvrb} % extended verbatim environments
%  \fvset{fontsize=\normalsize}% default font size for fancy-verbatim environments
\makeatletter
% Paragraph indentation and separation for normal text
\renewcommand{\@tufte@reset@par}{%
  \setlength{\RaggedRightParindent}{1.0pc}%
  \setlength{\JustifyingParindent}{1.0pc}%
  \setlength{\parindent}{0pc}%
  \setlength{\parskip}{0.5\baselineskip}%
}
\@tufte@reset@par

% Paragraph indentation and separation for marginal text
\renewcommand{\@tufte@margin@par}{%
  \setlength{\RaggedRightParindent}{0pt}%
  \setlength{\JustifyingParindent}{0.5pc}%
  \setlength{\parindent}{0.5pc}%
  \setlength{\parskip}{0pt}%
}
\makeatother

% Set up the spacing using fontspec features
\renewcommand\allcapsspacing[1]{{\addfontfeatures{LetterSpace=15}#1}}
\renewcommand\smallcapsspacing[1]{{\addfontfeatures{LetterSpace=10}#1}}


\title{Study Guide 15\hfill}
\author{Macroevolution}

\date{} % without \date command, current date is supplied

\begin{document}

\maketitle	% this prints the handout title, author, and date

\section{Vocabulary}\marginnote{\textbf{Read:} 515–516; 372–373 (heterotopy); 380–381 (related to Fig.~15.12).}

\begin{multicols}{2}
macroevolution \\
microevolution \\
heterotopy \\
bone morphogenetic protein\ 2\\\hspace{1em}(\textsc{bmp2})\\
calmodulin (\textsc{c}a\textsc{m1})
\textsc{bmp4}\\
sonic hedgehog (shh)\\

\end{multicols}

%\printclassoptions

\section{Concepts}

You should \emph{write} clear and concise answers\marginnote{This lecture ties macroevolutionary concepts to material covered in lecture 14 so I recommend you review it and its study guide again.} to each question in the Concepts section.  The questions are not necessarily independent.  Think broadly across lectures to see ``the big picture.'' 

\begin{enumerate}
	
	\item Explain why macroevolution is a necessary outcome of microevolutionary processes after speciation has occurred?
	
	\item Describe the basic evolution of jaws.  Use the original model, not the latest evidence (as that is still being figured out).

	\item Name two toolkit genes that have been implicated in the evolution of jaws.  What evidence exists from studies of modern vertebrates to support the hypothesis that these toolkit genes could be partly responsible for the evolution of jaws.

	\item Discuss the evolutionary importance of co-option at the genetic level.  Use the evolution of tetrapod limbs as an example. 

	\item Describe how the expression of co-opted hox genes in the tetrapod limb consistent with their original expression along the anterior-posterior vertebrate axis.

	\item Scientists have long held that the evolution of limbs has occurred independently several times within the metazoan kingdom.  For example, there is no structural homology between the appendages of crabs and the appendages of birds. Yet, analyses of the toolkit genes responsible for appendage development suggests homology at the genetic level.  Discuss the evidence that suggests the development of metazoan appendages is genetically homologous.  Be sure to name the genes (acronyms are fine, e.g., shh) and describe and illustrate the general patterns of expression for the genes in the developing limb.
	
	\item What toolkit genes have been implicated in the evolution of archosaurian scales\marginnote{You may have to look up archosaurs.} and bird feathers?  Which toolkit genes potentially determine the types of feathers found in birds?  What are the implications of this evidence for the evolution of mammalian hair? 

	\item Thought question. If birds evolved from dinosaurs, should feathered dinosaurs molt?

	\item Based on what you know about dinosaur-bird ancestry, construct a scientific answer for this question: Which came first, the chicken (bird) or the egg?
	
\end{enumerate}

\end{document}