These questions are based on Carroll, S.B., 1995. Homeotic genes and the
evolution of arthropods and chordates. Nature 376: 479-485.

Read the paper carefully and then type your answers. Hand-written
assignments will not be accepted. You must have your completed answers
with you in class or your assignment will be considered late. Any
assignment e-mailed to me after the \emph{start} of class will be
considered late. Do not copy answers directly from the text (that would
be plagiarism). \emph{Interpret} what you read into your own words. Be
prepared to discuss your answers in small groups and as a class. Failure
to discuss these questions may result in a pop quiz with a point value
equal to this assignment.

Failure to attend and participate in the in-class discussion on the due
date will result in an automatic 50\% deduction.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \begin{quote}
  What are the \emph{Hox} genes? How are they organized in the genome?
  How does this organization relate to their expression in the
  developing organism. Briefly describe the role of hox genes in the
  evolution of the arthropod and chordate body plans. (5 pts)
  \end{quote}
\item
  \begin{quote}
  Carroll states that \emph{Hox} proteins bind to a four nucleotide core
  sequence. Use the Internet to discover the exact sequence that is
  recognized by the \emph{Hox} proteins, and type it below (cite your
  source). Briefly but specifically describe how binding sites for
  \emph{Hox} proteins might be gained or lost by target genes. (5 pts)
  \end{quote}

\item
  \begin{quote}
  (5 pts)
  \end{quote}

\item
  \begin{quote}
  
  (5 pts)
  \end{quote}
\item
  \begin{quote}
   (5 pts)
  \end{quote}
\item
  \begin{quote}
  (5 pts).
  \end{quote}
\end{enumerate}
